Modeling the competing effects of the immune system and EMT on epithelial cancers

### 2019-09-16 Paper/Paper.tex
add line numbers
Rewrote after referees got back to us

\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}

\input{structure.tex} % Specifies the document structure and loads requires packages

@@

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}

%\usepackage[ ]{natbib} 
\newcommand{\tcr} { \textcolor{red} }

% Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer
\title{
Modeling the competing effects of the immune system and EMT on epithelial cancers
}

@@

\section*{Abstract}
Preceding and during cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor incidence and prognosis. Inflammation also has significant effects on gastrointestinal and pancreatic tumors. Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT). Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes. Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues. Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival. Delaying tumorigenesis depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness. We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions. We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer. These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.

%% Basic Introduction (1-2 sentences)
%Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
%% More Detailed Background (2-3 sentences)
%These are regulated by the tumor microenvironment (TME) in myriad ways. Moreover,  epithelial cell plasticity impacts disease incidence and progression via processes of epithelial-to-mesenchymal transition (EMT).
%In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
%% General Problem (1 sentence)
%Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
%% Summarizing Main Result (1 sentence)
%Here, we seek to understand how interplay between the tumor, the immune system, and EMT can determine the outcome of cancer.
%% The Main Result (2-3 sentences)
%We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
%The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival. In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis. Moreover, encouraging
%\tcr
%the TME
%towards a pro-EMT state also benefits the tumor.
%% Putting into a More General Context (1-2 sentences)
%Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.

%%%% Possible additional refs:
% Diefenbach 2001
% They show that when tumor cells express a particular marker, then the immune system gains considerable potency and clears the tumor. Furthermore, immunization via tumor cells expressing these ligands creates a host immunity to subsequent tumors even if these tumors lack the ligand.

% Carretero 2015
% As eosinophils can better carry out their recruitment function of CD8+ T cells, tumor rejection by the immune system is more readily accomplished.

% Ju 2016
% A cancer vaccine was demonstrated to be effective in reducing colorectal cancer incidence in mice. In addition, efficacy of CTLs was shown to be increased.

@@

Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}. The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
%Make sure there's some ref for immune clearing tumor
% Also, add that as immune cells clear tumor cells, they become deactivated and give source
In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}. Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}. 

@@

Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}. Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}. Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}. 

@@

While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.  TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}. Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

%Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} (which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below).

@@

Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT. We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when. A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT. We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.

@@

In the next section we develop the model and give explanation of the intuition behind each component. We then go on to analyze its behavior. Global sensitivity analysis identifies parameters that are crucial for tumorigenesis. We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival. To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.

%%%%%%%%%%%%%  INTRO PARAGRAPH GRAVEYARD %%%%%%%%%%%%%%%%

%They release cytokines that alter the extracellular matrix, promote angiogenesis, and create an inflammatory environment.
%This last one, in particular, results in the impairment of cytotoxic capabilities of immune cells.
%Through 
%
% its presence and by the cellular processes it engages in that result in shaping the TME.
%On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
%It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
%On the other hand, the tumor cells actively work to shape the immune response that the body launches.
%This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells. Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.

%In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
%This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
%For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''

%This plays an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
%As mesenchymal cells often exhibit stem-like properties, their existence in a tumor has been tied to the initiation and progression of cancer\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references.
%Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}. 
%
%Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell\cite{terry2017new}.
%Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
%Each of these three pieces will be briefly explored in turn for how they will influence one another.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.
\par
We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the TME, i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear mutated cells, the increased immune evasion only benefits mutated mesenchymal cells.

\subsection{Model Simulation}

\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the weight of a cell to proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be mutated ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

\paragraph{Rest in $G_0$} 
The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

@@

The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

%$$ \text{TGFB}_i = H(\text{TGFB}/N) + X_i, \quad X_i \sim N(0,\sigma^2)$$
\begin{equation}\tag{2.5}
\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{1}{1+K_3/\tau}+ X_i
, \quad X_i \sim N(0,\sigma^2)
\end{equation}

The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

\begin{equation}\tag{2.9}
\begin{split}
N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
\end{split}
\end{equation}

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}
\usepackage{lineno}

\input{structure.tex}

@@

\usepackage[normalem]{ulem}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}

\newcommand{\tcr} { \textcolor{red} }
\newcommand{\tcb} { \textcolor{blue} }

\title{
Modeling the competing effects of the immune system and EMT on epithelial cancers
}

@@

\section*{Abstract}
\sout{Preceding and d}\tcb{D}uring cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor \sout{incidence and} prognosis.
Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
Delaying \sout{tumorigenesis} \tcb{progression} depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
These results offer novel means to delay disease \sout{onset} \tcb{progression} by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.

\linenumbers

@@

Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
% Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}
% Similar concern to above as Guo's paper dealt with initiation, not progression.
\tcr{Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.}

@@

Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}.
Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}. 

@@

% Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor \tcr{initiation} \tcb{\cite{puisieux2014oncogenic}} and progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}. 
\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} \sout{(which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below)}.

@@

Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.
A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
\tcb{
In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
}
We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.

@@

In the next section we develop the model and give explanation of the intuition behind each component.
We then go on to analyze its behavior.
Global sensitivity analysis identifies parameters that are crucial for \sout{tumorigenesis} \tcb{progression}.
We then study these in more depth, focusing on the competing effects of EMT and of the immune system on \sout{tumorigenesis} \tcb{progression}: we find that EMT intricately regulates \sout{tumorigenesis} \tcb{progression} and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
\sout{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.}
\tcb{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.}
% Consider rewriting above

@@

\begin{table}[H]
\begin{tabular}{|p{\textwidth}|}
\hline
%%%
\tcb{
\begin{itemize}
\item TGF-$\beta$ is well-mixed within the TME
\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
\item All tissue cells (mutated and non-mutated) are equally likely to encounter immune cells
\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
\end{itemize}
}
\\
\hline
\end{tabular}
\caption{\tcb{Basic assumptions of the model}}
\label{table:model_assumptions}
\end{table}

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
\tcb{
In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
}
If the pathway is mutated, the component is colored red and is marked with `X'.
\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
\tcb{
% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
}
\par
We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
\tcb{
effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
}
The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
\tcb{
also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
}
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
% Above line replaced
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis. Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\tcb{
\subsection{Inflammation} % How did we not have a section on this before?
We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
For the purpose of the model, we consider low inflammation to be the default state.
When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
\tcr{See Table 2 in the supplement to see which parameters are changed by the inflammation score changing.} % Table 2 or S2? Change label in supplement?
}

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
\par
\tcb{
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
}
\subsection{Model Simulation}

\subsubsection{Initial conditions} 
Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
A set of warmup cycles are run to ensure arrival to this steady state.
During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
\sout{Thus, the only immune cell present is NKs.}
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the weight of a cell to proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
% Add: 
\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

\paragraph{Rest in $G_0$} 
\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by}
\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
Hence, the weight of rest is given by}
% Replace quiescence with rest and explain more above

@@

\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
\par
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common. So, this is not really so much a TGFB received as it is a change in TGFB signaling.
\sout{Each cell receives an amount of TGF-$\beta$ given by}
\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}

\begin{equation}\tag{2.5}
\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
, \quad X_i \sim N(0,\sigma^2)
\end{equation}

\tcb{
We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
}
% Need to add better explanation for Hill fn
\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
}
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

\begin{equation}\tag{2.9}
\begin{split}
N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
\end{split}
\end{equation}
